Ornaments as indicators of social changes in northeastern Taiwan before and after the European colonial period

### 2020-07-02 analysis/paper/paper.Rmd
edits based on R2 comments

Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade network could go beyond the temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often go through the processes of negotiations, resistance, or transformation that a static time frame may not capture the processes well [@Scheiber2012]. Also, we acknowledged that a fine-grained identification of decades of the European presences from the archaeological record is challenging in our research based on some reasons. The current radiocarbon ages that are unable to provide a high resolution of time scale, the circulation of foreign goods might start earlier before the European presences, and the use of the foreign goods in Indigenous societies could last longer after the official end of European colonization [@Mitchell2010]. It is more suitable to use pericolonial approaches that consider a long-term indirect consequences of foreign impacts rather than a focus on specific time frame [@Trabert2018; @Scheiber2012].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan. Frequency represents artifact counts", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories 
ornaments_count <-
  ornaments_period %>% 
  filter(Categories != "Unknown Metal") %>%
  mutate(Categories = ifelse(Categories == "Agate bead", "Carnelian bead", Categories),
         Categories = ifelse(Categories == "Golden bead", "Gold-foil bead", Categories),
         Categories = ifelse(Categories == "Bell", "Metal bell", Categories)) %>%
  ungroup() %>% # need to upgroup because of rowwise or factor in previous chunk?
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  group_by(Categories) %>%
  count() %>% 
  ungroup() %>% 
  mutate(Categories = fct_reorder(Categories, n))
plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)
ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.jpg"),
       width = 90, 
       height = 40,
       dpi = 300,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.jpg"))
```

@@

Archaeological contexts show that ornaments are especially abundant in burial contexts serving as grave goods [@Chen2007]. This supports the interpretation of ornaments as valuable objects functioning as status indicators. Spatial patterns of ornaments in dwelling contexts show that their distribution was clustered during the pre-European and European periods. These clusters are non-random, and are most highly concentrated during the European period. This may indicate that a degree of social inequality based on the uneven distribution of ornaments was already present before European contact, and then it was reinforced and amplified during the European period. A further indicator of increased social inequality is a burial dated to the 17th century that included 60 gold-foil beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008].

@@

The Kiwulan ornament data may be interpreted as indicating that Yilan social organization moved from a corporate mode, before the European arrival, to a network mode during European presence. The changes from a less concentrated to a more concentrated distribution of ornaments before and after the presence of Europeans appears consistent with the shift from shared power and wealth to accumulated wealth and monopolization. One possible explanation for the shift could be the long-distance trade network introduced by Europeans. The rarity and the image of colonial power of foreign trade goods resulted in the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting them [@Boone1992; @Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012]. 
That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms, but we have no specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.

>>>>

Despite the relatively short period of the European colonization in Taiwan (38-years-long Dutch occupation and 16-years-long Spanish occupation), the indirect impacts of the Europeans through trade network could go beyond the temporal boundaries and last longer. This is because the use of foreign goods in many Indigenous societies often go through the processes of negotiations, resistance, or transformation that a static time frame may not capture the processes well [@Scheiber2012]. Also, we acknowledged that a fine-grained identification of decades of the European presences from the archaeological record is challenging in our research based on some reasons. The current radiocarbon ages that are unable to provide a high resolution of time scale, the circulation of foreign goods might start earlier before the European presences, and the use of the foreign goods in Indigenous societies could last longer after the official end of European colonization [@Mitchell2010]. It is more suitable to use pericolonial approaches that consider indirect and wide-ranging effects of the colonial process such as trading activities [@Trabert2018; @Scheiber2012].

@@

```{r plot-all-ornaments-count, fig.cap= "Frequency of the major class of ornaments at Kiwulan. Frequency represents artifact counts totalizaing three periods", fig.width=image_width_1_col_inch}
# need to reorder, try https://forcats.tidyverse.org/reference/index.html
# reorder factor for categories (by frequency) and period (manually)
# total number of ornaments for each categories 
ornaments_count <-
  ornaments_period %>% 
  filter(Categories != "Unknown Metal") %>%
  mutate(Categories = ifelse(Categories == "Agate bead", "Carnelian bead", Categories),
         Categories = ifelse(Categories == "Golden bead", "Gold-foil bead", Categories),
         Categories = ifelse(Categories == "Bell", "Metal bell", Categories)) %>%
  ungroup() %>% # need to upgroup because of rowwise or factor in previous chunk?
  mutate(Categories = fct_lump(Categories, 
                               n = 5)) %>% 
  group_by(Categories) %>%
  count() %>% 
  ungroup() %>% 
  mutate(Categories = fct_reorder(Categories, n))
plot_ornaments_count <- 
ggplot(ornaments_count,
       aes(Categories, n)) +
  geom_col() +
  coord_flip() +
  scale_y_continuous(limits = c(0, 160)) +
  labs(y = "Frequency",
       x = "Categories") +
  theme_minimal(base_size = 10)
ggsave(plot = plot_ornaments_count,
       here("analysis", "figures",
       "plot-ornaments-count.jpg"),
       width = 90, 
       height = 40,
       dpi = 300,
       units = "mm")
knitr::include_graphics(here("analysis", 
                             "figures", 
                             "plot-ornaments-count.jpg"))
```

@@

Archaeological contexts show that ornaments are especially abundant in burial contexts serving as grave goods [@Chen2007]. This supports the interpretation of ornaments as valuable objects functioning as status indicators. Spatial patterns of ornaments in dwelling contexts show that their distribution was clustered during the pre-European and European periods. These clusters are non-random, and are most highly concentrated during the European period. This may indicate that a degree of social inequality based on the uneven distribution of ornaments was already present before European contact, and then it was reinforced and amplified during the European period. A further indicator of increased social inequality is a burial dated to the 17th century that included 60 gold-foil beads, well above the average of 2-3 pieces in the pre-European period [@Chen2007; @Cheng2008]. @Cheng2008 proposed a more hierarchical structure of Kiwulan based on the found of burial with 60 gold-foil beads that were treated as prestige goods in local community. However, @Hsieh's research on data from burial contexts suggested a more egalitarian society since burials with rich goods are elders that indicates accumulated wealth over time instead of inherited. Our results focusing on the residential area demonstrated an uneven distribution of prestige ornaments during the European presence that inclined an appearance of social inequality. Nevertheless, we still need a comprehensive exploration of other archaeological records to discuss the social structure of Kiwulan.

@@

The Kiwulan ornament data may be interpreted as indicating that Yilan social organization moved from a corporate mode, before the European arrival, to a network mode during European presence. The changes from a less concentrated to a more concentrated distribution of ornaments before and after the presence of Europeans appears consistent with the shift from shared power and wealth to accumulated wealth and monopolization. One possible explanation for the shift could be the long-distance trade network introduced by Europeans. The rarity and the image of colonial power of foreign trade goods resulted in the emergence of competition among ambitious individuals for prestige, wealth, or power through collecting them [@Boone1992; @Brumfiel1994; @Clark1994]. Because of the weak direct control from the European colonizers in northeastern Taiwan, local leaders may have had the flexibility to manipulate European colonial images, expand personal power, and monopolize the high-value trade goods [@Kang2012].

That said, the evidence from Kiwulan may be consistent with a variety of scenarios of indigenous-colonial relations. The increasing number and concentrated spatial patterns of ornaments may also indicate a practice of cultural resistance against the European intrusion. Resistance to European economic and political demands may be inferred if ornaments were used as a display of social identity and to emphasize the local customs that had existed before European contact [cf. @Rubertone2000]. Resistance could be presented in many forms, but we have no specific evidence from Kiwulan to prefer resistance as the primary mechanism behind the distribution of ornaments over colonial influence. Another scenario is that ornaments were treated as heirlooms, such as carnelian beads and gold-foil beads, that passed from one generation to the next, accumulating at Kiwulan over time. This process would result in a natural increase in ornaments over time, unrelated to colonial influences. As noted above, this pattern is not a good fit for the ornament distribution at Kiwulan, and adds little value in explaining the shifts in spatial patterns.
